Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-05-31 manuscript/main.tex
final submitted version of manuscript
including images and rendered PDF's

% Limit 350 words
\begin{abstract}
Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses. This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability. Here we explore this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites. We built process based phenology models for species common to each dataset. From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data. We found that model parameter estimates for the same species are most similar between the two datasets when using simple models, but parameter estimates vary widely as model complexity increases. Despite this, estimates for the date of phenological events and out-of-sample errors are similar, regardless of the model chosen. Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built. Accordingly, the choice of dataset depends on the research question. Inferences about species-specific phenological requirements are best made with LTER data, and if NPN or similar data are all that is available, then analyses should be limited to simple models. Large-scale predictive modeling is best done with the larger-scale NPN data, which has high spatial representation and a large regional species pool. LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well. Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
\end{abstract}

@@

Many plant phenology studies use intensively collected datasets from a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}. These datasets have regular sampling and large numbers of samples over long periods of time. As a result, the biological and climatic variability at that site is well represented. It is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age or population density \citep{kramer1995, diez2012}. For these reasons, data from a single location is not adequate for larger scale phenology modeling. Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.

@@

The National Phenology Network protocol uses status-based monitoring, where via an phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table S1).

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature. The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 ($NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}. These models are some of the most commonly used in phenology research and serve as a suitable baseline for comparing the NPN and LTER datasets.

@@

Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}. Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models. For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four LTER sites and across all NPN sites.

We also directly evaluated model performance using four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations. Within each of these scenarios, we calculated the RMSE of the held-out observations for each species, phenophase, and model type. Using the RMSE values, we calculated two different metrics to compare the performance of LTER and NPN-derived models on different data types. The first metric focuses on local-scale prediction by comparing the fits of LTER and NPN-derived models on LTER observations: $RMSE_{A} - RMSE_{B}$. The second metric focuses on large-scale prediction by comparing the fits of LTER and NPN-derived models on the NPN data: $RMSE_{C} - RMSE_{D}$. Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison (scenario A) within each site, not across sites, to focus on local scale prediction when LTER data is available. Absolute RMSE values are provided in the supplement (Fig. S1-S3).

We performed all analysis using both the R and Python programming languages \citep{rcitation, python}. Primary R packages used in the analysis included dplyr \citep{dplyr}, tidyr \citep{tidyr}, ggplot2 \citep{ggplot2}, lubridate \citep{lubridate}, prism \citep{prismR}, raster \citep{rasterR}, and sp \citep{sp1}. Primary Python packages included SciPy \citep{scipy}, NumPy \citep{numpy}, Pandas \citep{pandas}, and MPI for Python \citep{mpi4py}. Code to fully reproduce this analysis is available on GitHub (https://github.com/sdtaylor/phenology\_dataset\_study) and archived on Zenodo (XXXX)

@@

When comparing estimates of phenological events between the two sets of models, many NPN and LTER models produced similar estimates (Fig. 3). The \textit{Fixed GDD model} had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively). Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}. The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at NPN sites. The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) did not improve the correlation between datasets over the \textit{GDD model}.

@@

Despite substantial differences in parameter estimates, LTER and NPN-derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 produce nearly identical results in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Supplementary Material). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Supplementary Material). These results are consistent with research showing that models estimating the starting day of warming accumulation from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the Fixed GDD model which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets. In combination with this previous research, our results warrant caution in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the Fixed GDD model, based on their ability to make out-of-sample predictions. Based on the RMSE, the GDD and Uniforc models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig. S1 \& S2). This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}. In combination with our results showing observed differences in parameter estimates between LTS sites (Supplementary Material), this suggests that variation in phenological requirements across the the range is likely important. However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses. Specifically, correspondence between parameter estimates (Fig. 2), estimates of phenological events (Fig. 3), and out-of-sample error rates (Fig. 4) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively. This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the M1 model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by having a 40-year time-series across a large region. \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research. Large-scale phenology datasets, like NPN, will be key in addressing this and other phenological research needs.

In conclusion, our results suggest that both LTER and NPN data provide valuable information on plant phenology. Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events. The primary difference in the datasets is spatial scale, but due to tradeoffs in data collection efforts, the larger scale NPN data has shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3). These differences can be strengths or potential limitations. Observers sampling opportunistically allows the NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}. Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits its use in species-level predictive modeling. Due to these differences, the best data source for making predictions depends on the scale at which the predictions are being made. Identifying the most effective data sources for different types and scales of analysis is a useful first step, but the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}. Our results suggest that methods that can learn from the intensive information available in LTER data in regions where it is available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology. Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{richardson2018}. Data integration provides the potential to use data from many sources to produce the best opportunity for accurate inference about, and forecasting of, the timing of biological events.

@@

\begin{adjustbox}{angle=90}
%%renewcommand{\arraystretch}{2}\tabcolsep=5pt
    \begin{tabular}{ | l | l | l | l |}
    \hline
    Dataset Name & Habitat &  Phenological Event\newline (Num. Species) & Reference \\ \hline
    Harvard Forest & N.E. Deciduous\newline Forest & Budburst (16)\newline Flowering (15) & \citep{okeefe2015} \\
    Jornada Experimental Range & Chihuahuan Desert & Flowering (2) &  \\
    H.J. Andrews Experimental Forest & N.W. Wet Coniferous \newline Forest & Budburst (6)\newline Flowering (6) & \citep{schulze2017} \\
    Hubbard Brook & N.E. Deciduous \newline Forest & Budburst (3) & \citep{bailey2018} \\
    \hline
    \end{tabular}
\end{adjustbox} \newline
\textbf{Table 1}: LTER datasets used in the analysis

@@

\textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTS derived models. Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrap iterations. The black line is the 1:1 line. The $R^2$ is the coefficient of determination, which can be negative if the relationship between the two parameter sets is worse than no relationship but with the same mean values.

@@

\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between NPN and LTS-derived models. Top panels show comparisons at LTS sites and bottom panels show comparisons at NPN sites. Each point is an estimate for a single held-out observation. Colors indicate observations for a single species and phenophase combination.

@@

\textbf{Figure 4}: Differences in prediction error between NPN and LTS-derived models. Density plots for comparisons of predictions on LTS data (top row) and NPN data (bottom row). Each plot represents the difference between the RMSE for LTS-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTS-derived models and values greater than zero indicate more accurate prediction by NPN-derived models. Differences are calculated pairwise for the 38 species/phenophase comparisons.

>>>>

% Limit 350 words
\begin{abstract}
Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses. This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability. Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites. We built process based phenology models for species common to each dataset. From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data. We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased. Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen. Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built. Accordingly, the choice of dataset depends on the research question. Inferences about species-specific phenological requirements are best made with LTER data, and if NPN or similar data are all that is available, then analyses should be limited to simple models. Large-scale predictive modeling is best done with the larger-scale NPN data, which has high spatial representation and a large regional species pool. LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well. Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
\end{abstract}

@@

Many plant phenology studies use intensively collected datasets from a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}. These datasets have regular sampling and large numbers of samples over long periods of time. As a result, the biological and climatic variability at that site is well represented. It is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age or population density \citep{kramer1995, diez2012}. For these reasons, data from a single location is not adequate for larger scale phenology modeling. Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, tang2016, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.

@@

The National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table 1 \& S1).

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature. The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 ($NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}. These models are some of the most commonly used in phenology research and serve as a suitable baseline for comparing the NPN and LTER datasets.

@@

Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}. Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and NPN derived models for all held out observations. For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and NPN-derived estimates for estimates made at the four LTER sites and across all NPN sites.

We also directly evaluated model performance using four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations. Within each of these scenarios, we calculated the RMSE of the held-out observations for each species, phenophase, and model type. Using the RMSE values, we calculated two different metrics to compare the performance of LTER and NPN-derived models on different data types. The first metric focuses on local-scale prediction by comparing the fits of LTER and NPN-derived models on LTER observations: $RMSE_{A} - RMSE_{B}$. The second metric focuses on large-scale prediction by comparing the fits of LTER and NPN-derived models on the NPN data: $RMSE_{C} - RMSE_{D}$. These metrics were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison (scenario A) within each site, not across sites, to focus on local scale prediction when LTER data is available. Absolute RMSE values are provided in the supplement (Fig. S1-S3).

We performed all analysis using both the R and Python programming languages \citep{rcitation, python}. Primary R packages used in the analysis included dplyr \citep{dplyr}, tidyr \citep{tidyr}, ggplot2 \citep{ggplot2}, lubridate \citep{lubridate}, prism \citep{prismR}, raster \citep{rasterR}, and sp \citep{sp1}. Primary Python packages included SciPy \citep{scipy}, NumPy \citep{numpy}, Pandas \citep{pandas}, and MPI for Python \citep{mpi4py}. Code to fully reproduce this analysis is available on GitHub  (https://github.com/sdtaylor/phenology\_dataset\_study) and archived on Zenodo  (https://doi.org/10.5281/zenodo.1256705)

@@

When comparing estimates of phenological events between the two sets of models, many NPN and LTER models produced similar estimates (Fig. 3). The \textit{Fixed GDD model} had the highest correlation between the two models sets at NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively). Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}. The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at NPN sites. The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) improved the correlation over the \textit{GDD model} only slightly at LTER sites (from 0.51 to 0.52) and did not improve the correlation at NPN sites.

@@

Despite substantial differences in parameter estimates, LTER and NPN-derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T^*$=0 and $t_{1}$=5, $F$=5, $T^*$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Fig. S6-S7). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S4-S5). These results are consistent with research showing that models estimating the starting day of warming accumulation from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets. In combination with this previous research, our results warrant caution in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions. Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both NPN and LTER datasets (Fig. S1 \& S2). This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}. In combination with our results showing observed differences in parameter estimates between LTER sites (Fig. S4-S5), this suggests that variation in phenological requirements across the the range is likely important. However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses. Specifically, correspondence between parameter estimates (Fig. 2), estimates of phenological events (Fig. 3), and out-of-sample error rates (Fig. 4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively. This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by having a 40-year time-series across a large region. \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research. Large-scale phenology datasets, like NPN, will be key in addressing this and other phenological research needs.

In conclusion, our results suggest that both LTER and NPN data provide valuable information on plant phenology. Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events. The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale NPN data has shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3). These differences can be strengths or potential limitations. Observers sampling opportunistically allows the NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}. Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits its use in species-level predictive modeling. Due to these differences, the best data source for making predictions depends on the scale at which the predictions are being made. Identifying the most effective data sources for different types and scales of analysis is a useful first step, but the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}. Our results suggest that methods that can learn from the intensive information available in LTER data in regions where it is available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology. Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{richardson2018}. Data integration provides the potential to use data from many sources to produce the best opportunity for accurate inference about, and forecasting of, the timing of biological events.

@@

\begin{adjustbox}{angle=90}
%%renewcommand{\arraystretch}{2}\tabcolsep=5pt
    \begin{tabular}{ | l | l | l | l |}
    \hline
    Dataset Name & Habitat &  Phenological Event (Num. Species) & Reference \\ \hline
    Harvard Forest & N.E. Deciduous Forest & Budburst (17) Flowering (7) & \citep{okeefe2015} \\
    Jornada Experimental Range & Chihuahuan Desert & Flowering (2) &  \\
    H.J. Andrews Experimental Forest & N.W. Wet Coniferous Forest & Budburst (5) Flowering (4) & \citep{schulze2017} \\
    Hubbard Brook & N.E. Deciduous Forest & Budburst (3) & \citep{bailey2018} \\
    \hline
    \end{tabular}
\end{adjustbox} \newline
\textbf{Table 1}: LTER datasets used in the analysis

@@

\textbf{Figure 2}: Comparisons of parameter estimates between NPN and LTER derived models. Each point represents a parameter value for a specific species and phenophase, and is the mean value from 250 bootstrap iterations. The black line is the 1:1 line. The $R^2$ is the coefficient of determination, which can be negative if the relationship between the two parameter sets is worse than no relationship but with the same mean values.

@@

\textbf{Figure 3}: Comparison of predicted day of year (DOY) of all phenological events between NPN and LTER-derived models. Top panels show comparisons at LTER sites and bottom panels show comparisons at NPN sites. Each point is an estimate for a single held-out observation. Colors indicate observations for a single species and phenophase combination.

@@

\textbf{Figure 4}: Differences in prediction error between NPN and LTER-derived models. Density plots for comparisons of predictions on LTER data (top row) and NPN data (bottom row). Each plot represents the difference between the RMSE for LTER-derived model and the NPN-derived model, meaning that values less than zero indicate more accurate prediction by LTER-derived models and values greater than zero indicate more accurate prediction by NPN-derived models. Differences are calculated pairwise for the 38 species/phenophase comparisons.
